Semi-parametric generalized linear model for correlated binary data

### 2021-08-29 nuweb/SPGLM.tex
Insist on full range of probabilities

\begin{flushleft} \small\label{scrap1}\raggedright\small
\NWtarget{nuweb1}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {1}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@#' Fit semi-parametric GLM@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#'@{\tt @}\verb@rdname spglm@\\
\mbox{}\verb@#'@{\tt @}\verb@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.@\\
\mbox{}\verb@#'@{\tt @}\verb@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}@\\
\mbox{}\verb@#'@{\tt @}\verb@param subset  an optional vector specifying a subset of observations to be used.@\\
\mbox{}\verb@#'@{\tt @}\verb@param weights an optional vector specifying observation weights.@\\
\mbox{}\verb@#'@{\tt @}\verb@param offset  an optional vector specifying an offset term (predictor with a fixed coefficient of 1).@\\
\mbox{}\verb@#'@{\tt @}\verb@param link      a link function for the mean.@\\
\mbox{}\verb@#'@{\tt @}\verb@param mu0       an optional numeric value constraining the mean of the baseline distribution@\\
\mbox{}\verb@#'@{\tt @}\verb@param control a list with parameters controlling the algorithm.@\\
\mbox{}\verb@#'@{\tt @}\verb@return an object of class \code{spglm} with the fitted model.@\\
\mbox{}\verb@#'@{\tt @}\verb@export@\\
\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix model.frame model.offset model.response model.weights@\\
\mbox{}\verb@@\\
\mbox{}\verb@spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL, @\\
\mbox{}\verb@                  control=list(eps=0.001, maxit=100)){@\\
\mbox{}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb2}{2}}$\,\rangle$}\verb@@\\
\mbox{}\verb@    @\\
\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N, spt=0:N)@\\
\mbox{}\verb@    @\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb7}{7}}$\,\rangle$}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@    mt <- attr(mf, "terms")@\\
\mbox{}\verb@    names(betas) <- colnames(mm)@\\
\mbox{}\verb@    names(referencef0) <- 0:N@\\
\mbox{}\verb@    rownames(vc) <- colnames(vc) <- c(names(betas), names(referencef0))@\\
\mbox{}\verb@    res <- list(coefficients = betas, f0=referencef0, vcov = vc, mu0=mu0, niter = iter, @\\
\mbox{}\verb@                loglik=llik, link = link, call = mc, terms = mt,@\\
\mbox{}\verb@                xlevels = .getXlevels(mt, mf),@\\
\mbox{}\verb@                data_object=data_object)@\\
\mbox{}\verb@    class(res) <- "spglm"@\\
\mbox{}\verb@    res@\\
\mbox{}\verb@@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.

@@

\begin{flushleft} \small\label{scrap8}\raggedright\small
\NWtarget{nuweb8}{} $\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize {8}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@  # replace each cluster with N+1 clusters of size N@\\
\mbox{}\verb@  new <- cbind(y=0:N)@\\
\mbox{}\verb@  rep_idx <- rep(1:nrow(Y), each=nrow(new))@\\
\mbox{}\verb@  Y2 <- cbind(1:nrow(Y), Y)[rep_idx,]@\\
\mbox{}\verb@  colnames(Y2) <- c("i", "resp","nonresp")@\\
\mbox{}\verb@@\\
\mbox{}\verb@  rep_idx2 <- rep(1:nrow(new), times=nrow(Y))@\\
\mbox{}\verb@  Ycomb <- cbind(Y2, new[rep_idx2, ,drop=FALSE])@\\
\mbox{}\verb@  # select possible combinations@\\
\mbox{}\verb@  possible <- (Ycomb[,"y"] >= Ycomb[,"resp"]) & (N-Ycomb[,"y"] >= Ycomb[,"nonresp"])@\\
\mbox{}\verb@  Ycomb <- Ycomb[possible,]@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  obs_start <- match(1:nobs, Ycomb[,"i"])@\\
\mbox{}\verb@ @\\
\mbox{}\verb@  mm2 <- mm[Ycomb[,"i"], ,drop=FALSE]@\\
\mbox{}\verb@  weights2 <- weights[Ycomb[,"i"]]@\\
\mbox{}\verb@  offset2 <- offset[Ycomb[,"i"]]@\\
\mbox{}\verb@@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.

@@

\begin{flushleft} \small\label{scrap12}\raggedright\small
\NWtarget{nuweb10b}{} $\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize {10b}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # identify possible y-values@\\
\mbox{}\verb@  spt <- sort(unique(Ycomb[,"y"]))@\\
\mbox{}\verb@  data_object$spt <- spt@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # ensure all positive values within support@\\
\mbox{}\verb@  referencef0[spt+1] <- referencef0[spt+1] + 1e-6@\\
\mbox{}\verb@  referencef0 <- referencef0 / sum(referencef0)@\\
\mbox{}\verb@@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  fTiltMatrix <- matrix(rep(referencef0, times=nobs), byrow=TRUE, @\\
\mbox{}\verb@      nrow=nobs, ncol=N+1)@\\
\mbox{}\verb@@\\
\mbox{}\verb@  if (is.null(mu0))@\\
\mbox{}\verb@    mu0 <- weighted.mean(Y[,1]/rowSums(Y), weights)@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # hypergeometric terms for log-likelihood calculation@\\
\mbox{}\verb@  hp <- sapply(0:N, function(t)dhyper(x=Y[,1], m=rowSums(Y), n=N-rowSums(Y), k=t))   @\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # initial log-likelihood@\\
\mbox{}\verb@  llik <- log(rowSums(fTiltMatrix * hp)) %*% weights@\\
\mbox{}\verb@@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.

@@

\begin{flushleft} \small\label{scrap13}\raggedright\small
\NWtarget{nuweb11}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {11}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@  ll <- function(x){@\\
\mbox{}\verb@    spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object=data_object, link=link)@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@@\\
\mbox{}\verb@  hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))@\\
\mbox{}\verb@  hess_idx <- c(1:p, p+spt+1) # betas and f0 elements with non-zero support@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # revert to unlogged f0@\\
\mbox{}\verb@  grad <- c(rep(1, p), 1/referencef0[spt+1])@\\
\mbox{}\verb@  hess1 <- diag(grad) %*% hess[hess_idx, hess_idx] %*% diag(grad)@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # create bordered hessian@\\
\mbox{}\verb@  border1 <- c(rep(0, p), rep(1, length(spt)))  # gradient of sum-to-one constraint@\\
\mbox{}\verb@  border2 <- c(rep(0, p), spt)          # gradient of fixed-mean constraint@\\
\mbox{}\verb@  bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # calculate variance-covariance matrix@\\
\mbox{}\verb@  bvc <- solve(-bhess)@\\
\mbox{}\verb@  # remove border and set to 0 for zero-support values@\\
\mbox{}\verb@  vc <- matrix(0, nrow=p+N+1, ncol=p+N+1)@\\
\mbox{}\verb@  vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     @\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.

>>>>

\begin{flushleft} \small\label{scrap1}\raggedright\small
\NWtarget{nuweb1}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {1}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@#' Fit semi-parametric GLM@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#'@{\tt @}\verb@rdname spglm@\\
\mbox{}\verb@#'@{\tt @}\verb@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.@\\
\mbox{}\verb@#'@{\tt @}\verb@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}@\\
\mbox{}\verb@#'@{\tt @}\verb@param subset  an optional vector specifying a subset of observations to be used.@\\
\mbox{}\verb@#'@{\tt @}\verb@param weights an optional vector specifying observation weights.@\\
\mbox{}\verb@#'@{\tt @}\verb@param offset  an optional vector specifying an offset term (predictor with a fixed coefficient of 1).@\\
\mbox{}\verb@#'@{\tt @}\verb@param link      a link function for the mean.@\\
\mbox{}\verb@#'@{\tt @}\verb@param mu0       an optional numeric value constraining the mean of the baseline distribution@\\
\mbox{}\verb@#'@{\tt @}\verb@param control a list with parameters controlling the algorithm.@\\
\mbox{}\verb@#'@{\tt @}\verb@return an object of class \code{spglm} with the fitted model.@\\
\mbox{}\verb@#'@{\tt @}\verb@export@\\
\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix model.frame model.offset model.response model.weights@\\
\mbox{}\verb@@\\
\mbox{}\verb@spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL, @\\
\mbox{}\verb@                  control=list(eps=0.001, maxit=100)){@\\
\mbox{}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb2}{2}}$\,\rangle$}\verb@@\\
\mbox{}\verb@    @\\
\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, @\\
\mbox{}\verb@                        maxN=N, spt=0:N)@\\
\mbox{}\verb@    @\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb7}{7}}$\,\rangle$}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@    mt <- attr(mf, "terms")@\\
\mbox{}\verb@    names(betas) <- colnames(mm)@\\
\mbox{}\verb@    names(referencef0) <- 0:N@\\
\mbox{}\verb@    rownames(vc) <- colnames(vc) <- c(names(betas), names(referencef0))@\\
\mbox{}\verb@    res <- list(coefficients = betas, f0=referencef0, vcov = vc, mu0=mu0, niter = iter, @\\
\mbox{}\verb@                loglik=llik, link = link, call = mc, terms = mt,@\\
\mbox{}\verb@                xlevels = .getXlevels(mt, mf),@\\
\mbox{}\verb@                data_object=data_object)@\\
\mbox{}\verb@    class(res) <- "spglm"@\\
\mbox{}\verb@    res@\\
\mbox{}\verb@@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.

@@

To ensure that the range of possible response probabilities covers 0 -- 1, we need to ensure that clusters with 0 and $N$ responses both are possible, buy adding small-weight fake clusters if needed.

\begin{flushleft} \small\label{scrap8}\raggedright\small
\NWtarget{nuweb8}{} $\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize {8}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # identify if fake observations are needed @\\
\mbox{}\verb@  # we need at least one observation with 0 events and one with 0 non-events@\\
\mbox{}\verb@  fakeID <- numeric()@\\
\mbox{}\verb@  # save original values@\\
\mbox{}\verb@  Y0 <- Y@\\
\mbox{}\verb@  mm0 <- mm@\\
\mbox{}\verb@  offset0 <- offset@\\
\mbox{}\verb@  weights0 <- weights@\\
\mbox{}\verb@  nobs0 <- nobs@\\
\mbox{}\verb@ @\\
\mbox{}\verb@  if (!(0 %in% Y[,1])){  # 0 events@\\
\mbox{}\verb@    Y <- rbind(Y, c(0, N))@\\
\mbox{}\verb@    mm <- rbind(mm, colMeans(mm0))@\\
\mbox{}\verb@    offset <- c(offset, mean(offset0))@\\
\mbox{}\verb@    weights <- c(weights, 1e-7)@\\
\mbox{}\verb@    nobs <- nobs + 1@\\
\mbox{}\verb@    fakeID <- c(fakeID, nobs)@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@  if (!(0 %in% Y[,2])){ # 0 non-events@\\
\mbox{}\verb@    Y <- rbind(Y, c(N, 0))@\\
\mbox{}\verb@    mm <- rbind(mm, colMeans(mm0))@\\
\mbox{}\verb@    offset <- c(offset, mean(offset0))@\\
\mbox{}\verb@    weights <- c(weights, 1e-7)@\\
\mbox{}\verb@    nobs <- nobs + 1@\\
\mbox{}\verb@    fakeID <- c(fakeID, nobs)@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # replace each cluster with N+1 clusters of size N@\\
\mbox{}\verb@  new <- cbind(y=0:N)@\\
\mbox{}\verb@  rep_idx <- rep(1:nrow(Y), each=nrow(new))@\\
\mbox{}\verb@  Y2 <- cbind(1:nrow(Y), Y)[rep_idx,]@\\
\mbox{}\verb@  colnames(Y2) <- c("i", "resp","nonresp")@\\
\mbox{}\verb@@\\
\mbox{}\verb@  rep_idx2 <- rep(1:nrow(new), times=nrow(Y))@\\
\mbox{}\verb@  Ycomb <- cbind(Y2, new[rep_idx2, ,drop=FALSE])@\\
\mbox{}\verb@  # select possible combinations@\\
\mbox{}\verb@  possible <- (Ycomb[,"y"] >= Ycomb[,"resp"]) & (N-Ycomb[,"y"] >= Ycomb[,"nonresp"])@\\
\mbox{}\verb@  Ycomb <- Ycomb[possible,]@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  obs_start <- match(1:nobs, Ycomb[,"i"])@\\
\mbox{}\verb@ @\\
\mbox{}\verb@  mm2 <- mm[Ycomb[,"i"], ,drop=FALSE]@\\
\mbox{}\verb@  weights2 <- weights[Ycomb[,"i"]]@\\
\mbox{}\verb@  offset2 <- offset[Ycomb[,"i"]]@\\
\mbox{}\verb@@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.

@@

\begin{flushleft} \small\label{scrap12}\raggedright\small
\NWtarget{nuweb10b}{} $\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize {10b}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # identify possible y-values@\\
\mbox{}\verb@  spt <- sort(unique(Ycomb[,"y"]))@\\
\mbox{}\verb@  data_object$spt <- spt@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  #real_spt <- sort(unique(Ycomb[!(Ycomb[,"i"] %in% fakeID),"y"]))@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # ensure all positive values within support@\\
\mbox{}\verb@  referencef0[spt+1] <- referencef0[spt+1] + 1e-6@\\
\mbox{}\verb@  referencef0 <- referencef0 / sum(referencef0)@\\
\mbox{}\verb@@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  fTiltMatrix <- matrix(rep(referencef0, times=nobs), byrow=TRUE, @\\
\mbox{}\verb@      nrow=nobs, ncol=N+1)@\\
\mbox{}\verb@@\\
\mbox{}\verb@  if (is.null(mu0))@\\
\mbox{}\verb@    mu0 <- weighted.mean(Y[,1]/rowSums(Y), weights)@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # hypergeometric terms for log-likelihood calculation@\\
\mbox{}\verb@  hp <- sapply(0:N, function(t)dhyper(x=Y[,1], m=rowSums(Y), n=N-rowSums(Y), k=t))   @\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # initial log-likelihood@\\
\mbox{}\verb@  llik <- log(rowSums(fTiltMatrix * hp)) %*% weights@\\
\mbox{}\verb@@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.

@@

\begin{flushleft} \small\label{scrap13}\raggedright\small
\NWtarget{nuweb11}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {11}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@  data_object_mod <- list(model_matrix = mm, resp=Y, n=rowSums(Y), @\\
\mbox{}\verb@                          weights = weights, offset = offset, maxN = N, @\\
\mbox{}\verb@                          spt = spt)@\\
\mbox{}\verb@  ll <- function(x){@\\
\mbox{}\verb@    spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object=data_object_mod, link=link)@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@@\\
\mbox{}\verb@  hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))@\\
\mbox{}\verb@  hess_idx <- c(1:p, p+spt+1) # betas and f0 elements with non-zero real support@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # revert to unlogged f0@\\
\mbox{}\verb@  grad <- c(rep(1, p), 1/referencef0[spt+1])@\\
\mbox{}\verb@  hess1 <- diag(grad) %*% hess[hess_idx, hess_idx] %*% diag(grad)@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # create bordered hessian@\\
\mbox{}\verb@  border1 <- c(rep(0, p), rep(1, length(spt)))  # gradient of sum-to-one constraint@\\
\mbox{}\verb@  border2 <- c(rep(0, p), spt)          # gradient of fixed-mean constraint@\\
\mbox{}\verb@  bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # calculate variance-covariance matrix@\\
\mbox{}\verb@  bvc <- solve(-bhess)@\\
\mbox{}\verb@  # remove border and set to 0 for zero-support values@\\
\mbox{}\verb@  vc <- matrix(0, nrow=p+N+1, ncol=p+N+1)@\\
\mbox{}\verb@  vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     @\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.
